---
title: Probability Cheatsheet
format: html
---

```{r}
#| output: false
#| message: false
#| warning: false
#| echo: false

library(eulerr)
```

## Complement of an Event

In general, for a given event $A$, the complement is the subset of other outcomes that do not belong to event $A$:

$$1 = P(A) + P(A^c),$$ 

where $^c$ means the complement (we read it as "not").

## Conditional Independence

Random variables $X$ and $Y$ are conditionally independent given random variable $Z$ if and only if

$$P(X = x \cap Y = y \mid Z = z) = P(X = x \mid Z = z) \cdot P(Y = y \mid Z = z).$$

## Conditional Probability

In general, let $A$ and $B$ be two events of interest within the sample $S$, and $P(B) > 0$, then the conditional probability of $A$ given $B$ is defined as:

$$P(A \mid B) = \frac{P(A \cap B)}{P(B)}$$

Note event $B$ is becoming the new sample space (i.e., $P(B \mid B) = 1$). The tweak here is that our original sample space $S$ has been updated to $B$. 

## Covariance

Let $X$ and $Y$ be two numeric random variables; their covariance is defined as follows:

$$\operatorname{Cov}(X, Y) = \mathbb{E}[(X-\mu_X)(Y-\mu_Y)],$$

where $\mu_X = \mathbb{E}(X)$ and $\mu_Y = \mathbb{E}(Y)$ are the respective means (or expected values) of $X$ and $Y$. After some algebraic and expected value manipulations, the above equation reduces to a more practical form to work with:

$$\operatorname{Cov}(X,Y) = \mathbb{E}(XY) - \left[ \mathbb{E}(X)\mathbb{E}(Y) \right],$$

where $\mathbb{E}(XY)$ is the mean (or expected value) of the multiplication of the random variables $X$ and $Y$.

## Cumulative Distribution Function

Let $X$ be a continuous random variable with probability density function (PDF) $f_X(x)$. The cumulative distribution function (CDF) is usually denoted by $F(\cdot)$ and is defined as 

$$F_X(x) = P(X \leq x).$$

We can calculate the CDF by 

$$F_X(x) = \int_{-\infty}^x f_X(t) \, \text{d}t.$$

In order for $F_X(x)$ to be a valid CDF, the function needs to satisfy the following requirements:

1. Must never decrease.
2. It must never evalute to be $< 0$ or $> 1$. 
3. $F_X(x) \rightarrow 0$ as $x \rightarrow -\infty$
4. $F_X(x) \rightarrow 1$ as $x \rightarrow \infty$.

## Entropy

Let $X$ be a random variable:

- If $X$ is discrete, with $P(X = x)$ as a probability mass function (PMF), then the entropy is defined as:

$$H(Y) = -\displaystyle \sum_x P(X = x)\log[P(X = x)].$$

- If $X$ is continuous, with $f_X(x)$ as a probability density function (PDF), then the entropy is defined as: 

$$H(X) = -\int_x f_X(x) \log [f_X(x)] \text{d}x.$$

Note that, in Statistics, the $\log(\cdot)$ notation implicates base $e$.

## Expected Value

Let $X$ be a numeric random variable. The mean $\mathbb{E}(X)$ (also known as expected value or expectation) is defined as:

- If $X$ is discrete, with $P(X = x)$ as a probability mass function (PMF), then

$$\mathbb{E}(X) = \displaystyle \sum_x x \cdot P(X = x).$$

- If $X$ is continuous, with $f_X(x)$ as a probability density function (PDF), then 

$$\mathbb{E}(X) = \displaystyle \int_x x \cdot f_X(x) \text{d}x.$$

In general for a function of $X$ such as $g(X)$, the expected value is defined as:

- If $X$ is discrete, with $P(X = x)$ as a PMF, then

$$\mathbb{E}\left[ g(X) \right] = \displaystyle \sum_x g(X) \cdot P(X = x).$$

- If $X$ is continuous, with $f_X(x)$ as a PDF, then 

$$\mathbb{E}\left[ g(X) \right] = \displaystyle \int_x g(X) \cdot f_X(x) \text{d}x.$$

## Inclusion-Exclusion Principle

### Two Events

Let $A$ and $B$ be two events of interest in the sample space $S$. The probability of $A$ or $B$ occuring is denoted as $P(A \cup B)$, where $\cup$ means **"OR."** The Inclusion-Exclusion Principle allows us to compute this probability as:

$$P(A \cup B) = P(A) + P(B) - P(A \cap B),$$

where $P(A \cap B)$ denotes the probability of $A$ and $B$ occuring simultaneously ($\cap$ means **"AND"**). 

$P(A \cup B)$ can be represented with the overall shaded area in the Venn diagram depicted in @fig-union-A-B.

```{r}
#| label: fig-union-A-B
#| fig-cap: "Venn diagram depicting the union of events A and B."
#| echo: false 
#| message: false
#| warning: false
#| fig.height: 4
#| fig.width: 6

AB_venn <- euler(c("A" = 0.75, "B" = 0.75, "A&B" = 0.1))

plot(AB_venn,
  quantities = FALSE,
  fills = c("rosybrown1", "lightsteelblue1"),
  labels = list(cex = 1.2),
  main = "\n Sample Space S"
)
```

### Three Events

We can also extend this principle to three events ($A,$ $B$, and $C$ in the sample space $S$):

$$
\begin{align*}
P(A \cup B \cup C) &= P(A) + P(B) + P(C) - \\
& \qquad P(A \cap B) - P(B \cap C) - P(A \cap C) + \\
& \qquad \quad P(A \cap B \cap C),
\end{align*}
$$

where $P(A \cap B \cap C)$ denotes the probability of $A$, $B$, and $C$ occuring simultaneously.

$P(A \cup B \cup C)$ can be represented with the overall shaded area in the Venn diagram depicted in @fig-union-A-B-C

```{r}
#| label: fig-union-A-B-C
#| fig-cap: "Venn diagram depicting the union of events A, B, and C."
#| echo: false 
#| message: false
#| warning: false
#| fig.height: 4
#| fig.width: 6

ABC_venn <- euler(c(
  "A" = 0.75, "B" = 0.75, "C" = 0.75,
  "A&B" = 0.1, "A&C" = 0.1, "B&C" = 0.1,
  "A&B&C" = 0.1
))

plot(ABC_venn,
  quantities = FALSE, 
  fills = c("rosybrown1", "lightsteelblue1", "moccasin"),
  labels = list(cex = 1.2),
  main = "\n Sample Space S"
)
```

## Independent Events

Let $A$ and $B$ be two events of interest in the sample space $S$. These two events are independent if the occurrence of one of them does not affect the probability of the other. In probability notation, their intersection is defined as:

$$P(A \cap B) = P(A) \cdot P(B).$$

## Independence in Probability Distributions between Two Random Variables

Let $X$ and $Y$ be two independent random variables. Using their corresponding marginals, we can obtain their corresponding joint distributions as follows:

- **$X$ and $Y$ are discrete.** Let $P(X = x, Y = y)$ be the joint probability mass function (PMF) with $P(X = x)$ and $P(Y = y)$ as their marginals. Then, we define the joint PMF as:
  
$$P(X = x, Y = y) = P(X = x) \cdot P(Y = y).$$

The term denoting a discrete joint PMF $P(X = x, Y = y)$ is equivalent to the intersection of events $P(X = x \cap Y = y)$.

- **$X$ and $Y$ are continuous.** Let $f_{X,Y}(x,y)$ be the joint probability density function (PDF) with $f_X(x)$ and $f_Y(y)$ as their marginals. Then, we define the joint PDF as:
  
$$f_{X,Y}(x,y) = f_X(x) \cdot f_Y(y).$$

## Independent Random Variables

Let $X$ and $Y$ be two random variables. We say $X$ and $Y$ are independent if knowing something about one of them tells us nothing about the other. A definition of $X$ and $Y$ being independent is the following:

$$P(X = x \cap Y = y) = P(X = x) \cdot P(Y = y).$$

## Kendall's $\tau_K$

Let $X$ and $Y$ be two numeric random variables. Kendall's $\tau_K$ measures concordance between each pair of observations $(x_i, y_i)$ and $(x_j, y_j)$ with $i \neq j$:

- **Concordant**, which gets a positive sign, means 
$$\begin{gather*}
x_i < x_j \quad \text{and} \quad y_i < y_j, \\
\text{or} \\
x_i > x_j \quad \text{and} \quad y_i > y_j.
\end{gather*}$$

- **Discordant**, which gets a negative sign, means
$$\begin{gather*}
x_i < x_j \quad \text{and} \quad y_i > y_j, \\
\text{or} \\
x_i > x_j \quad \text{and} \quad y_i < y_j.
\end{gather*}$$

Mathematically, we can set it up as:

$$\tau_K = \frac{\text{Number of concordant pairs} - \text{Number of discordant pairs}}{{n \choose 2}},$$

with the "true" Kendall's $\tau_K$ value obtained by sending $n \rightarrow \infty$. Here, $n$ is the sample size (i.e., the number of data points). Note that:

$$-1 \leq \tau_K \leq 1.$$

## Law of Total Expectation

Let $X$ and $Y$ be two numeric random variables. Generally, a marginal mean $\mathbb{E}_Y(Y)$ can be computed from the conditional means $\mathbb{E}_Y(Y \mid X = x)$ and the probabilities of the conditioning variables $P(X = x)$:

$$\mathbb{E}_Y(Y) = \sum_x \mathbb{E}_Y(Y \mid X = x) \cdot P(X = x).$$

Or, it can also be written as:

$$\mathbb{E}_Y(Y) = \mathbb{E}_X [\mathbb{E}_Y(Y \mid X)].$$

Also, the previous result extends to probabilities:

$$P(Y = y \cap X = x) = P(Y = y \mid X = x) \cdot P(X = x).$$

## Linearity of Expectations

If $a$ and $b$ are constants, with $X$ and $Y$ as numeric random variables, then we can obtain the expected value of the following expressions as:

$$\begin{gather*}
\mathbb{E}(a X) = a \mathbb{E}(X) \\
\mathbb{E}(X + Y) = \mathbb{E}(X) + \mathbb{E}(Y) \\
\mathbb{E}(aX + bY) = a\mathbb{E}(X) + b\mathbb{E}(Y).
\end{gather*}$$

## Linearity of Variances with Two Independent Random Variables

If $a$ and $b$ are constants, with $X$ and $Y$ as independent numeric random variables, then we can obtain the variance of the following expressions as:

$$\begin{gather*}
\text{Var}(a X) = a^2 \text{Var}(X) \\
\text{Var}(X + Y) = \text{Var}(X) + \text{Var}(Y) \\
\text{Var}(aX + bY) = a^2 \text{Var}(X) + b^2 \text{Var}(Y).
\end{gather*}$$

## Marginal (Unconditional) Probability

In general, the probability of an event $A$ occurring is denoted as $P(A)$ and is defined as 

$$P(A) = \frac{\text{Number of times event $A$ is observed}}{\text{Total number of events observed}}.$$

## Median

Let $X$ be a numeric random variable. The median $\text{M}(X)$ is the outcome for which there is a 50-50 chance of seeing a greater or lesser value. So, its distribution-based definition satisfies 

$$P[X \leq \text{M}(X)] = 0.5.$$

## Mode

Let $X$ be a random variable:

- If $X$ is discrete, with $P(X = x)$ as a probability mass function (PMF), then the mode is the outcome having the highest probability.
- If $X$ is continuous, with $f_X(x)$ as a probability density function (PDF), then the mode is the outcome having the highest density. That is:

$$\text{Mode} = {\arg \max}_x f_X(x).$$

## Mutual Information

The mutual information between two random variables $X$ and $Y$ is defined as

$$H(X,Y) = \displaystyle \sum_x \displaystyle \sum_y P(X = x \cap Y = y)\log\left[\frac{P(X = x \cap Y = y)}{P(X = x) \cdot P(Y = y)}\right].$$

## Mutually Exclusive (or Disjoint) Events

Let $A$ and $B$ be two events of interest in the sample space $S$. These events are mutually exclusive (or disjoint) if they cannot happen at the same time in the sample space $S$. Thus, in probability notation, their intersection will be:

$$P(A \cap B) = 0.$$

Therefore, by the Inclusion-Exclusion Principle, the union of these two events can be obtained as follows:

$$\begin{align*}
P(A \cup B) &= P(A) + P(B) - \underbrace{P(A \cap B)}_{0} \\
&= P(A) + P(B).
\end{align*}$$

These two events are shown in Venn diagram depicted in @fig-disjoint-A-B.

```{r}
#| label: fig-disjoint-A-B
#| fig-cap: "Venn diagram depicting two mutually exclusive (or disjoint) events A and B."
#| echo: false 
#| message: false
#| warning: false
#| fig.height: 4
#| fig.width: 6

AB_disj_venn <- euler(c("A" = 0.75, "B" = 0.75))

plot(AB_disj_venn,
  quantities = FALSE,
  fills = c("rosybrown1", "lightsteelblue1"),
  labels = list(cex = 1.2),
  main = "\n Sample Space S"
)
```

## Odds

Let $p$ be the probability of an event of interest $A$. The odds $o$ is the ratio of the probability of the event $A$ to the probability of the non-event $A$:

$$o = \frac{p}{1 - p}.$$

In plain words, the odds will tell how many times event $A$ is more likely compared to how unlikely it is.

## Pearson's Correlation

Let $X$ and $Y$ be two numeric random variables, whose respective variances are defined by the variance equation, with a covariance defined as above. Pearson's correlation standardizes the distances according to the standard deviations $\sigma_X$ and $\sigma_Y$ of $X$ and $Y$, respectively. It is defined as:

$$\begin{align*}
\rho_{XY} = \operatorname{Corr}(X, Y) &= \mathbb{E} \left[ 
   \left(\frac{X-\mu_X}{\sigma_X}\right) 
   \left(\frac{Y-\mu_Y}{\sigma_Y}\right)
 \right] \\
 &= \frac{\operatorname{Cov}(X, Y)}{\sqrt{\operatorname{Var}(X)\operatorname{Var}(Y)}}.
\end{align*}$$

As a result of the above equation, it turns out that

$$-1 \leq \rho_{XY} \leq 1.$$

## Probability of a Continuous Random Varible $X$ Being between $a$ and $b$

For a continuous random variable $X$ with probability density function (PDF) $f_X(x)$, the probability of $X$ being between $a$ and $b$ is 

$$P(a \leq X \leq b) = \int_a^b f_X(x)\text{d}x.$$

We can connect the dots with our new definition of a cumulative distribution function (CDF). First,

$$P(a \leq X \leq b) = P(X \leq b) - P(X \leq a)$$

because if $X \leq b$ but not $\leq a$ then it must be that $a \leq X \leq b$. But now we can write these two terms using the CDF:

$$P(a \leq X \leq b) = P(X \leq b) - P(X \leq a) = F_X(b) - F_X(a).$$

Now, plugging in the definition of the CDF as the integral of the PDF,

$$P(a \leq X \leq b) = \int_{-\infty}^b f_X(x) \, \text{d}x - \int_{-\infty}^a f_X(x) \, \text{d}x=\int_{a}^b f_X(x) \, \text{d}x.$$

## Properties of the Bivariate Gaussian or Normal Distribution

Let $X$ and $Y$ be part of a bivariate Gaussian or Normal distribution with means $-\infty < \mu_X < \infty$ and $-\infty < \mu_Y < \infty$, variances $\sigma^2_X > 0$ and $\sigma^2_Y > 0$, and correlation coefficient $-1 \leq \rho_{XY} \leq 1$.

This bivariate Gaussian or Normal distribution has the following properties:

1. **Marginal distributions are Gaussian.** The marginal distribution of a subset of variables can be obtained by just taking the relevant subset of means, and the relevant subset of the covariance matrix.

2. **Linear combinations are Gaussian.** This is actually by definition. If $(X, Y)$ have a bivariate Gaussian or Normal distribution with marginal means $\mu_X$ and $\mu_Y$ along with marginal variances $\sigma^2_X$ and $\sigma^2_Y$ and covariance $\sigma_{XY}$; then $Z = aX + bY + c$ with constants $a, b, c$ is Gaussian. If we want to find the mean and variance of $Z$, we apply the linearity of expectations and variance rules:

$$
\begin{align*}
\mathbb{E}(Z) &= \mathbb{E}(aX + bY + c) \\
&= \mathbb{E}(aX) + \mathbb{E}(bY) + \mathbb{E}(c) \\
&= a \mathbb{E}(X) + b \mathbb{E}(Y) + c \\
&= a \mu_X + b \mu_Y + c \\
\\
\text{Var}(Z) &= \text{Var}(aX + bY + c) \\
&= \text{Var}(aX) + \text{Var}(bY) + \text{Var}(c) + 2 \text{Cov}(aX, bY) \\
&= a^2 \text{Var}(X) + b^2 \text{Var}(Y) + 0 + 2ab \text{Cov}(X, Y) \\
&= a^2 \sigma_X^2 + b^2 \sigma_Y^2 + 2ab \sigma_{XY}.
\end{align*}
$$

3. **Conditional distributions are Gaussian.** If $(X, Y)$ have a bivariate Gaussian or Normal distribution with marginal means $\mu_X$ and $\mu_Y$ along with marginal variances $\sigma^2_X$ and $\sigma^2_Y$ and covariance $\sigma_{XY}$; then the distribution of $Y$ given that $X = x$ is also Gaussian. Its distribution is

$$Y \mid X = x \sim \mathcal{N} \left(\mu_{_{Y \mid X = x}} = \mu_Y + \frac{\sigma_Y}{\sigma_X} \rho_{XY} (x - \mu_X), \sigma^2_{_{Y \mid X = x}} = \ (1 - \rho_{XY}^2)\sigma_Y^2 \right).$$

## Quantile

Let $X$ be a numeric random variable. A $p$-quantile $Q(p)$ is the outcome with a probability $p$ of getting a smaller outcome. So, its distribution-based definition satisfies 

$$P[X \leq Q(p)] = p.$$

## Quantile Function

Let $X$ be a continuous random variable. The quantile function $Q(\cdot)$ takes a probability $p$ and maps it to the $p$-quantile. It turns out that this is the inverse of the cumulative distribution function (CDF):

$$Q(p) = F^{-1}(p).$$

Note that this function does not exist outside of $0 \leq p \leq 1$. This is unlike the other functions (density, CDF, and survival function) which exist on all real numbers.

## Skewness

Let $X$ be a numeric random variable: 

- If $X$ is discrete, with $P(X = x)$ as a probability mass function (PMF), then skewness can be defined as

$$\text{Skewness}(X) = \mathbb{E} \left[ \left( \frac{X - \mu_X}{\sigma_X} \right)^3 \right] = \displaystyle \sum_x \left( \frac{x - \mu_X}{\sigma_X} \right)^3 \cdot P(X = x).$$ 

- If $X$ is continuous, with $f_X(x)$ as a probability density function (PDF), then 

$$\text{Skewness}(X) = \mathbb{E} \left[ \left( \frac{X - \mu_X}{\sigma_X} \right)^3 \right] = \displaystyle \int_x \left( \frac{x - \mu_X}{\sigma_X} \right)^3 \cdot f_X(x) \text{d}x.$$

where $\mu_X = \mathbb{E}(X)$ and $\sigma_X = \text{SD}(X)$.

## Survival Function

Let $X$ be a continuous random variable. The survival function $S(\cdot)$ is the cumulative distribution function (CDF) "*flipped upside down*". For this random variable $X$, the survival function is defined as

$$S_X(x) = P(X > x) = 1 - F_X(x).$$

## Variance

Let $X$ be a numeric random variable. The variance, either for a discrete or continuous random variable, is defined as 

$$
\begin{align*}
\text{Var}(X) &= \mathbb{E}\{[X - \mathbb{E}(X)]^2\}\\
&= \mathbb{E}(X^2) - [\mathbb{E}(X)]^2.
\end{align*}
$$

For the continuous case with $f_X(x)$ as a probability density function (PDF), an alternative definition of the variance is

$$\text{Var}(X) = \mathbb{E}[(X - \mu_X)^2] = \int_x (x - \mu_X) ^ 2 \, f_X(x) \text{d}x.$$

The term $\mu_X$ is equal to $\mathbb{E}(X)$ from the expected value equation.

Finally, either for a discrete or continuous random variable, the standard deviation is the square root of the variance:

$$\text{SD}\left[ \text{Var}(X) \right] = \sqrt{\text{Var}(X)}.$$

The above measure is more practical because it is on the same scale as the outcome, unlike the variance.

## Variance of a Sum Involving Two Non-Independent Random Variables

Suppose $X$ and $Y$ are not independent numeric random variables. Therefore, the variance of their sum is:

$$\operatorname{Var}(X + Y) = \operatorname{Var}(X) + \operatorname{Var}(Y) + 2\operatorname{Cov}(X, Y).$$

Furthermore, if $X$ and $Y$ are independent, then 

$$\mathbb{E}(XY) = \mathbb{E}(X) \mathbb{E}(Y).$$

Therefore, using the above equation, the sum becomes:

$$\operatorname{Var}(X + Y) = \operatorname{Var}(X) + \operatorname{Var}(Y).$$
